OptiFit: a fast method for fitting amplicon sequences to existing OTUs

### 2021-07-21 exploratory/2021/algorithm_diagram.Rmd

```{r ggraph_exp, fig.dim=c(6,5), dpi=300}
plot_graph <- function(graph, title = '') {
  create_layout(graph, 'linear', sort.by = id) %>% 
  ggraph() +
  geom_edge_arc(aes(label = mcc,
                    start_cap = label_rect(node1.name),
                    end_cap = label_rect(node2.name)), 
                arrow = arrow(length = unit(4, 'mm'),
                              type = 'closed'),
                edge_colour = 'gray',
                angle_calc = 'along',
                label_dodge = unit(-2, 'mm')
                ) +
  geom_edge_loop(aes(span = 1, direction = 90, strength = 0.2,# direction = 270, 
                     color = is_loop)) +
  geom_node_label(aes(label = name)) +
    scale_edge_color_manual(values = c('black')#c('white')
                            ) +
  labs(title = title) +
  theme_void() +
  theme(plot.margin=unit(x=c(0,0,10,0), units="pt"),
        legend.position = 'none')
}

@@

```{r plot_optifit_iters, fig.dim=c(6.875,9.0625), dpi=300}
# TODO custom modifications for loop span & direction
i <- 0
lapply(optifit_iters, function(x) {
  i <<- i + 1
  tbl_graph(nodes = x$nodes, edges = x$edges) %>% 
    plot_graph(title = glue('{i}) mcc = {x$edges %>% filter(is_loop) %>% pull(mcc)}'))
}) %>% 
  wrap_plots(ncol = 1)
```

>>>>

```{r ggraph_exp, fig.dim=c(6,5), dpi=300}
plot_graph <- function(graph, title = '', 
                       is_last_plot = FALSE,
                       hide_loops = FALSE) {
  loop_dir <- ifelse(is_last_plot, 270, 90)
  loop_color <- ifelse(hide_loops, 'white', 'black')
  create_layout(graph, 'linear', sort.by = id) %>% 
  ggraph() +
  geom_edge_arc(aes(label = mcc,
                    start_cap = label_rect(node1.name),
                    end_cap = label_rect(node2.name)), 
                arrow = arrow(length = unit(4, 'mm'),
                              type = 'closed'),
                edge_colour = 'gray',
                angle_calc = 'along',
                label_dodge = unit(-2, 'mm')
                ) +
  geom_edge_loop(aes(span = 1, 
                     direction = loop_dir, 
                     strength = 0.2,
                     color = is_loop)) +
  geom_node_label(aes(label = name)) +
    scale_edge_color_manual(values = c(loop_color)
                            ) +
  labs(title = title) +
  theme_void() +
  theme(plot.margin=unit(x=c(0,0,10,0), units="pt"),
        legend.position = 'none')
}

@@

```{r plot_optifit_iters, fig.dim=c(6.875,6.875), dpi=300}
i <- 0
lapply(optifit_iters, function(x) {
  i <<- i + 1
  tbl_graph(nodes = x$nodes, edges = x$edges) %>% 
    plot_graph(title = glue('{i}) mcc = {x$edges %>% filter(is_loop) %>% pull(mcc)}'),
               is_last_plot = i == length(optifit_iters),
               hide_loops = FALSE)
}) %>% 
  wrap_plots(ncol = 1)
```

### 2021-07-21 paper/paper.Rmd
Insert placeholder for algorithm diagram

OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
their sequences, a list of query sequences to assign to the reference OTUs, and
the sequence pairs that are within the distance threshold (e.g. 0.03).
Initially, all query sequences are placed into separate OTUs. 
Then, the algorithm iteratively reassigns the query sequences to the reference 
OTUs to optimize the MCC. 
Alternatively, a sequence will remain unassigned if the MCC value is maximized
when the sequence is a singleton rather than assigned to a reference OTU.
All query and reference sequence pairs are considered when calculating the MCC.
This process is repeated until the MCC changes by no more than 0.0001 (default)
or until a maximum number of iterations is reached (default: 100).
In the closed reference mode, any query sequences that cannot be assigned to
references OTUs are discarded, and the results will only contain OTUs that exist
in the original reference.
In the open reference mode, unassigned query sequences are clustered _de novo_
using OptiClust to generate new OTUs.
The final MCC is reported with the best OTU assignments.
There are two strategies for generating OTUs with OptiFit: 
1) fit the query sequences to reference OTUs generated by _de novo_ clustering an
independent database, or
2) split the dataset into a reference and query fraction, cluster the reference
sequences _de novo_, then fit the query sequences to the reference OTUs.
We clustered sequences from four datasets isolated from soil
[@johnston_metagenomics_2016], marine [@henson_artificial_2016], mouse gut
[@schloss_stabilization_2012], and human gut [@baxter_microbiota-based_2016]
samples to test the performance of OptiFit with both of these strategies.

@@

While _de novo_ clustering produces high quality OTUs, researchers may prefer to
perform reference clustering to a public database because reference-based
methods produce consistent OTUs and are generally faster than _de novo_ methods.
In closed reference mode, sequences that cannot be assigned to reference OTUs
are thrown out, so that the final clustering contains only OTUs that exist in
the reference.
To test how OptiFit performs for this purpose, we fit each dataset to three
databases of reference OTUs: the Greengenes database, the SILVA non-redundant
database, and the Ribosomal Database Project (RDP) [@desantis_greengenes_2006;
@quast_silva_2013; @cole_ribosomal_2014].
Reference OTUs for each database were created by performing _de novo_ clustering
with OptiClust at a distance threshold of 3% (see Figure \ref{fig:workflow}).
The _de novo_ MCC scores were `r dn_dbs_mcc` for `r db_names_upper`,
respectively.
Fitting sequences to Greengenes and SILVA in closed reference mode performed
similarly, with median MCC scores of `r round(closed_fit_gg_mcc, 3)` (**TODO: Pat marked this as 0.80, but the actual number is 0.795414**) and 
`r closed_fit_silva_mcc` respectively, while the median MCC 
was `r closed_fit_rdp_mcc` when fitting to RDP 
(see Figure \ref{fig:results_sum}).
For comparison, clustering datasets with OptiClust produced an average MCC score
of `r opticlust_mcc`.
This gap in OTU quality mostly disappeared when clustering in open reference
mode, which produced median MCCs of `r open_fit_gg_mcc` with Greengenes, 
`r open_fit_silva_mcc` with SILVA, and `r open_fit_rdp_mcc` with the RDP.
Thus, open reference OptiFit produced OTUs of very similar quality as _de novo_
clustering, and closed reference OptiFit followed closely behind as long as a
suitable reference database was chosen.

>>>>

```{r algorithm, fig.dim=figs('algorithm','dim'), fig.cap=figs('algorithm','cap')}
figs('algorithm', 'path') %>% 
  tiff::readTIFF() %>% 
  grid::grid.raster() 
```

OptiFit begins where OptiClust ends, starting with a list of reference OTUs and
their sequences, a list of query sequences to assign to the reference OTUs, and
the sequence pairs that are within the distance threshold (e.g. 0.03) 
(Figure \ref{fig:algorithm}).
Initially, all query sequences are placed into separate OTUs. 
Then, the algorithm iteratively reassigns the query sequences to the reference 
OTUs to optimize the MCC. 
Alternatively, a sequence will remain unassigned if the MCC value is maximized
when the sequence is a singleton rather than assigned to a reference OTU.
All query and reference sequence pairs are considered when calculating the MCC.
This process is repeated until the MCC changes by no more than 0.0001 (default)
or until a maximum number of iterations is reached (default: 100).
In the closed reference mode, any query sequences that cannot be assigned to
references OTUs are discarded, and the results will only contain OTUs that exist
in the original reference.
In the open reference mode, unassigned query sequences are clustered _de novo_
using OptiClust to generate new OTUs.
The final MCC is reported with the best OTU assignments.
There are two strategies for generating OTUs with OptiFit: 
1) fit the query sequences to reference OTUs generated by _de novo_ clustering an
independent database, or
2) split the dataset into a reference and query fraction, cluster the reference
sequences _de novo_, then fit the query sequences to the reference OTUs.
We clustered sequences from four datasets isolated from soil
[@johnston_metagenomics_2016], marine [@henson_artificial_2016], mouse gut
[@schloss_stabilization_2012], and human gut [@baxter_microbiota-based_2016]
samples to test the performance of OptiFit with both of these strategies.

@@

While _de novo_ clustering produces high quality OTUs, researchers may prefer to
perform reference clustering to a public database because reference-based
methods produce consistent OTUs and are generally faster than _de novo_ methods.
In closed reference mode, sequences that cannot be assigned to reference OTUs
are thrown out, so that the final clustering contains only OTUs that exist in
the reference.
To test how OptiFit performs for this purpose, we fit each dataset to three
databases of reference OTUs: the Greengenes database, the SILVA non-redundant
database, and the Ribosomal Database Project (RDP) [@desantis_greengenes_2006;
@quast_silva_2013; @cole_ribosomal_2014].
Reference OTUs for each database were created by performing _de novo_ clustering
with OptiClust at a distance threshold of 3% (see Figure \ref{fig:workflow}).
The _de novo_ MCC scores were `r dn_dbs_mcc` for `r db_names_upper`,
respectively.
Fitting sequences to Greengenes and SILVA in closed reference mode performed
similarly, with median MCC scores of `r closed_fit_gg_mcc` and 
`r closed_fit_silva_mcc` respectively, while the median MCC 
was `r closed_fit_rdp_mcc` when fitting to RDP 
(see Figure \ref{fig:results_sum}).
For comparison, clustering datasets with OptiClust produced an average MCC score
of `r opticlust_mcc`.
This gap in OTU quality mostly disappeared when clustering in open reference
mode, which produced median MCCs of `r open_fit_gg_mcc` with Greengenes, 
`r open_fit_silva_mcc` with SILVA, and `r open_fit_rdp_mcc` with the RDP.
Thus, open reference OptiFit produced OTUs of very similar quality as _de novo_
clustering, and closed reference OptiFit followed closely behind as long as a
suitable reference database was chosen.
